Title

### 2013-03-11 inc/model_results/shapes/shapes.tex
some initial updates on model results

The initial `space-time' eccentricity \ref{eq:define_xi}, $\xi ( 0 ) = v_{\smallzero} \tau / w $ , of the photo-generated electron pulse plays a major role in the bunch propagation dynamics, particularly when space-charge effects dominate.
\ref{fig:compare_shape} illustrates the typical dynamics of initially spherical ($ \sigma_{ \smallT }(0) = \sigma_{z} ( 0 ) $), `disk-like' ($ \sigma_{ \smallT } ( 0 ) > \sigma_{ z } ( 0 ) $), and `cigar-like' ($ \sigma_{ z } ( 0 ) > \sigma_{ \smallT } ( 0 ) $) pulses.
%TODO rewrite using w?
For simplicity, and unlike most simulations in this paper, the pulses in this simulation are not calculated from laser and photocathode parameters, but are rather defined directly.
The initial bunch dimensions are given by the ellipticities $ \xi(0) = \{ 1/3 , 1 , 3 \} $ while in each case the product of the initial HW1/eM bunch widths $w_{\smallT}^2 w_z = $ 1mm$^3$.
Since the initial pulse volumn is consistent between all runs, initial peak pulse charge density is six orders of magnitude greater in \ref{fig:compare_shape}\subref{fig:compare_shape_N1e6} ($ N = 10^{6} $ electrons in the bunch) than in \ref{fig:compare_shape}\subref{fig:compare_shape_N1} ($ N = 10^{0} $).
The pulse velocity is $ v_{\smallzero} \approx c/3 $, i.e. a DC photogun with $V = 30\text{kV} $.

\begin{figure}
  \centering
  \subfloat[][]{
    \label{fig:compare_shape_N1}
    \input{shapes_N1}
  }
  \\
  \subfloat[][]{
    \label{fig:compare_shape_N1e6}
    \input{shapes_N1e6}
  }
  \caption[AG simulation of free-space pulse expansion]{
    Transverse (solid) and longitudinal (dashed) dynamics of three example pulses for three length/width ratios ($\xi$). 
    The red lines represent a prolate pulse (a ``pancake'') with $\xi=1/3$, green lines for a spherical pulse with $\xi=1$, and blue for oblate (a ``cigar'') having $\xi=3$.
  }
  \label{fig:compare_shape}
\end{figure}

The bunch propagation dynamics for negligible pulse charge density ($ N = 10^{0} $) shown in \ref{fig:compare_shape}\subref{fig:compare_shape_N1} display the trends expected from the initial photoemission conditions.
In particular, in the initially spherical case ($ \sigma_{\smallT} (0) = \sigma_{z} (0) $), the transverse bunch width $ \sqrt{2 \sigma_{\smallT}} $ expands roughly twice as rapidly as the longitudinal length $ \sqrt{2 \sigma_{z}} $ due to the factor of two difference in the initial momentum variances $ \Delta p_{\alpha} $ and the artificially zero initial pulse chirp ($ \gamma_{\alpha} = 0 $).
Naturally, for initially non-spherical electron bunches, the pulse propagation model also shows that the relative bunch broadening $ \sqrt{ \sigma_{\alpha} / \sigma_{\alpha} (0) } $ for comparable $ \eta_{\alpha} $ is greatest for the smallest initial $ \sigma_{\alpha} (0) $; that is both disk- and cigar-like pulses are initially driven towards the spherical regime.
In all cases, this intrinsic electron pulse broadening due to non-zero $ \eta_{\alpha} $ generates a momentum chirp across the electron bunch; in the longitudinal dimension the faster electrons photoemitted with larger excess energy leading the slower electrons.

For an electron pulse with a higher initial space-charge density, intra-pulse Coulomb effects can dominate the propagation dynamics.
In \ref{fig:compare_shape}\subref{fig:compare_shape_N1e6}, we display the predicted temporal dynamics of electron pulses with the same initial size and photoemission conditions as \ref{fig:compare_shape}\subref{fig:compare_shape_N1}, but with $ N = 10^{6} $ electrons --- the number required for single-shot diffraction pattern measurements in a DTEM.\cite{berger_dc_2009,armstrong_practical_2007}
The significant influence of space-charge effects for the higher initial peak pulse charge density is clearly evident.
%TODO revisit this next sentance, the data doesn't seem to bear it out
In the spherical case, space-charge induced bunch broadening overwhelms the initial difference between $\eta_{\smallT}$ and $\eta_{z}$, producing an almost uniform three-dimensional `Coulomb explosion' of the electron pulse with a one-dimensional bunch broadening rate roughly one order of magnitude greater than that for $ N = 10^{0} $ (\ref{fig:compare_shape}\subref{fig:compare_shape_N1}).
For $ \xi (0) > 1 $ (or $ \xi (0) < 1 $), the smaller initial transverse (longitudinal) bunch dimension experiences the greatest relative expansion in free-space propagation. This is to be expected, because the initial $\eta_{\alpha}$ are non-zero (but comparable in magnitude) while Coulomb effects will produce differential pulse broadening in the dimension with the largest internal space-charge field (i.e., the dimension with the largest electrostatic potential gradient).
In fact, both the $ N = 10^{6} $ disk- and cigar-like pulse examples shown in \ref{fig:compare_shape}\subref{fig:compare_shape_N1e6} transition through a spherical bunch shape ($ \xi \approx 1 $) after a time of flight of $\sim$0.5ns, at which point the peak pulse charge density has already been reduced by a factor of $\sim$100.
The initial impulse induced by the strong space-charge field then provides for continued linear expansion (in time), resulting in a linear pulse chirp in the dimension associated with the smallest initial HW1/eM bunch width\cite{michalik_analytic_2006,berger_dc_2009} $ \sqrt{ 2 \sigma_{\alpha} (0) } $.

The data presented in \ref{fig:compare_shape} using the AG model of Michalik and Sipe \cite{michalik_analytic_2006} are entirely consistent with prior works.\cite{reed_femtosecond_2006,siwick_ultrafast_2002}
In femtosecond UED experiments, significant efforts are made to offset space-charge induced temporal (longitudinal, $z$) pulse broadening by reducing the time of flight from the gun photocathode to the sample.\cite{siwick_ultrafast_2002,reed_evolution_2009}
Moreover, efforts are now underway to compensate for temporal electron pulse broadening using RF cavity pulse compression techniques.\cite{oudheusden_electron_2007}
In UEM, employing a standard retro-fitted electron microscope column driven by MHz repetition-rate sub-picosecond lasers,\cite{lobastov_four-dimensional_2005} deleterious space-charge effects have obliged operation in the $\sim$1 electron/pulse regime to maintain high spatial resolution.
Even for nanosecond DTEM (the extreme cigar-like pulse ($\xi \gg 1 $)), transverse space-charge induced beam broadening has necessitated the insertion of an additional magnetic electron lens in a microscope column to allow more efficient delivery of electrons to the specimen when $ N > 10^{6} $.\cite{lagrange_nanosecond_2008}
In this paper, however, we will restrict the discussion to ultrashort (ps and sub-ps) electron pulses which (i) allow for the temporal resolution of fundamental events occurring on ultrafast time scales in materials science, biology and chemistry \cite{king_ultrafast_2005} and (ii) can be generated in the disk-like regime ($ \xi < 1 $) where transverse bunch broadening effects are minimized (\ref{fig:compare_shape})(enabling more efficient beam propagation) and RF cavities can be used to compensate and reverse longitudinal (temporal) pulse broadening.\cite{veisz_hybrid_2007}

I have shown that a prolate spheroid pulse shape is preferred for UEM, both for generation in SECTION %TODO link to Child law discussion
and in this section for propagation.
Further I have shown in %TODO link to section where I show that initial conditions create shorter pulses than laser
that electron pulses shorter than the laser pulse duration are likely occurances.
Henceforth I shall only consider prolate pulses when a choice of initial pulse shape is possible.

>>>>

The initial `space-time' eccentricity \ref{eq:define_xi}, $\xi ( 0 ) = v_{\smallzero} \tau / w $ , of the photo-generated electron pulse plays a major role in the bunch propagation dynamics, particularly when space-charge effects dominate.
\ref{fig:compare_shape} illustrates the typical dynamics of initially spherical ($ \sigma_{ \smallT }(0) = \sigma_{z} ( 0 ) $), `disk-like' ($ \sigma_{ \smallT } ( 0 ) > \sigma_{ z } ( 0 ) $), and `cigar-like' ($ \sigma_{ z } ( 0 ) > \sigma_{ \smallT } ( 0 ) $) pulses.
%TODO rewrite using w?
For simplicity, and unlike most simulations in this paper, the pulses in this simulation are not calculated from laser and photocathode parameters, but are rather defined directly.
%TODO should I remove this? Should I just specify the creation types on each instance?
The initial bunch dimensions are given by the ellipticities $ \xi(0) = \{ 1/3 , 1 , 3 \} $ while in each case the product of the initial HW1/eM bunch widths $w_{\smallT}^2 w_z = $ 1mm$^3$.
Since the initial pulse volume is consistent between all runs, initial peak pulse charge density is six orders of magnitude greater in \ref{fig:compare_shape}\subref{fig:compare_shape_N1e6} ($ N = 10^{6} $ electrons in the bunch) than in \ref{fig:compare_shape}\subref{fig:compare_shape_N1} ($ N = 10^{0} $).
The pulse velocity is $ v_{\smallzero} \approx c/3 $, i.e. a DC photogun with $V = 30\text{kV} $.

\begin{figure}
  \centering
  \subfloat[][]{
    \label{fig:compare_shape_N1}
    \input{shapes_N1}
  }
  \\
  \subfloat[][]{
    \label{fig:compare_shape_N1e6}
    \input{shapes_N1e6}
  }
  \caption[AG simulation of free-space pulse expansion]{
    Transverse (solid) and longitudinal (dashed) dynamics of three example pulses for three length/width ratios ($\xi$). 
    The red lines represent a prolate pulse (a ``pancake'') with $\xi=1/3$, green lines for a spherical pulse with $\xi=1$, and blue for an oblate pulse (a ``cigar'') having $\xi=3$.
    The pulses shown in \ref{fig:compare_shape_N1} all have total charge $N=1$, while in \ref{fig:compare_shape_N1e6} the pulses have $N=10^6$.
  }
  \label{fig:compare_shape}
\end{figure}

The bunch propagation dynamics for negligible pulse charge density ($ N = 10^{0} $) shown in \ref{fig:compare_shape}\subref{fig:compare_shape_N1} display the trends expected from the initial photoemission conditions.
In particular, in the initially spherical case ($ \sigma_{\smallT} (0) = \sigma_{z} (0) $), the transverse bunch width $ \sqrt{2 \sigma_{\smallT}} $ expands approximately twice as rapidly as the longitudinal length $ \sqrt{2 \sigma_{z}} $ due to the factor of two difference in the initial momentum variances $ \Delta p_{\alpha} $ and the artificially zero initial pulse chirp ($ \gamma_{\alpha} = 0 $).
Naturally, for initially non-spherical electron bunches, the pulse propagation model also shows that the relative bunch broadening $ \sqrt{ \sigma_{\alpha} / \sigma_{\alpha} (0) } $ for comparable $ \eta_{\alpha} $ is greatest for the smallest initial $ \sigma_{\alpha} (0) $; that is both disk- and cigar-like pulses are initially driven towards the spherical regime.
In all cases, this intrinsic electron pulse broadening due to non-zero $ \eta_{\alpha} $ generates a momentum chirp across the electron bunch; in the longitudinal dimension the faster electrons photoemitted with larger excess energy leading the slower electrons.

For an electron pulse with a higher initial space-charge density, intra-pulse Coulomb effects can dominate the propagation dynamics.
\ref{fig:compare_shape}\subref{fig:compare_shape_N1e6}, shows the predicted temporal dynamics of electron pulses with the same initial size and photoemission conditions as \ref{fig:compare_shape}\subref{fig:compare_shape_N1}, but with $ N = 10^{6} $ electrons --- the number required for single-shot diffraction pattern measurements in a DTEM.\cite{berger_dc_2009,armstrong_practical_2007}
The significant influence of space-charge effects for the higher initial peak pulse charge density is clearly evident.
%TODO revisit this next sentance, the data doesn't seem to bear it out
In the spherical case, space-charge induced bunch broadening overwhelms the initial difference between $\eta_{\smallT}$ and $\eta_{z}$, producing an almost uniform three-dimensional `Coulomb explosion' of the electron pulse with a one-dimensional bunch broadening rate roughly one order of magnitude greater than that for $ N = 10^{0} $ (\ref{fig:compare_shape}\subref{fig:compare_shape_N1}).
For $ \xi (0) > 1 $ (or $ \xi (0) < 1 $), the smaller initial transverse (longitudinal) bunch dimension experiences the greatest relative expansion in free-space propagation. This is to be expected, because the initial $\eta_{\alpha}$ are non-zero (but comparable in magnitude) while Coulomb effects will produce differential pulse broadening in the dimension with the largest internal space-charge field (i.e., the dimension with the largest electrostatic potential gradient).
In fact, both the $ N = 10^{6} $ disk- and cigar-like pulse examples shown in \ref{fig:compare_shape}\subref{fig:compare_shape_N1e6} transition through a spherical bunch shape ($ \xi \approx 1 $) after a time of flight of $\sim$0.5ns, at which point the peak pulse charge density has already been reduced by a factor of $\sim$100.
The initial impulse induced by the strong space-charge field then provides for continued linear expansion (in time), resulting in a linear pulse chirp in the dimension associated with the smallest initial HW1/eM bunch width\cite{michalik_analytic_2006,berger_dc_2009} $ \sqrt{ 2 \sigma_{\alpha} (0) } $.

The data presented in \ref{fig:compare_shape} using the AG model of Michalik and Sipe \cite{michalik_analytic_2006} are entirely consistent with prior works.\cite{reed_femtosecond_2006,siwick_ultrafast_2002}
In femtosecond UED experiments, significant efforts are made to offset space-charge induced temporal (longitudinal, $z$) pulse broadening by reducing the time of flight from the gun photocathode to the sample.\cite{siwick_ultrafast_2002,reed_evolution_2009}
Moreover, efforts are now underway to compensate for temporal electron pulse broadening using RF cavity pulse compression techniques.\cite{oudheusden_electron_2007}
In UEM, employing a standard retro-fitted electron microscope column driven by MHz repetition-rate sub-picosecond lasers,\cite{lobastov_four-dimensional_2005} deleterious space-charge effects have obliged operation in the $\sim$1 electron/pulse regime to maintain high spatial resolution.
Even for nanosecond DTEM (the extreme cigar-like pulse ($\xi \gg 1 $)), transverse space-charge induced beam broadening has necessitated the insertion of an additional magnetic electron lens in a microscope column to allow more efficient delivery of electrons to the specimen when $ N > 10^{6} $.\cite{lagrange_nanosecond_2008}
In this thesis, however, I will restrict the discussion to ultrashort (ps and sub-ps) electron pulses which (i) allow for the temporal resolution of fundamental events occurring on ultrafast time scales in materials science, biology and chemistry \cite{king_ultrafast_2005} and (ii) can be generated in the disk-like regime ($ \xi < 1 $) where transverse bunch broadening effects are minimized (\ref{fig:compare_shape})(enabling more efficient beam propagation) and RF cavities can be used to compensate and reverse longitudinal (temporal) pulse broadening.\cite{veisz_hybrid_2007}

%TODO perhaps all of this should be moved back or even cut?
I have shown that a prolate spheroid pulse shape is preferred for UEM, both for generation in SECTION %TODO link to Child law discussion
and in this section for propagation.
Further I have shown in %TODO link to section where I show that initial conditions create shorter pulses than laser
that electron pulses shorter than the laser pulse duration are likely occurances.
Henceforth I shall only consider prolate pulses when a choice of initial pulse shape is possible.

### 2013-03-11 main.tex
consistent use of v_max vs v_0
added magnetic lens loops comparison figure and moved some of the related content from the hardware section to the extension
some initial updates on model results

\providecommand{\smallT}{\ensuremath{{\scriptscriptstyle T}}}
\providecommand{\smallD}{\ensuremath{{\scriptscriptstyle D}}}
\providecommand{\smallF}[0]{ { \scriptscriptstyle F } }
%\providecommand{\smallZ}{\ensuremath{{\scriptscriptstyle Z}}}

@@

  \subimport{inc/extension/external_forces/}{external_forces}

@@

>>>>

\providecommand{\smallT}{\ensuremath{{\scriptscriptstyle T}}}
\providecommand{\smallD}{\ensuremath{{\scriptscriptstyle D}}}
\providecommand{\smallzero}{\ensuremath{{\scriptscriptstyle 0}}}
\providecommand{\smallF}[0]{ { \scriptscriptstyle F } }
%\providecommand{\smallZ}{\ensuremath{{\scriptscriptstyle Z}}}

@@

  \subimport{inc/extension/external_forces/}{formal}

  \subimport{inc/extension/external_forces/}{practical}

@@

In Chapters \ref{chap:previous_models} and \ref{chap:extension}, I have introduced the extended Analytic Gaussian model.
In the following chapter I use this model to make general observations about pulse propagation for Ultrafast Electron Microscopy.
